Mechanical loading prediction though accelerometry data during walking and running

### 2021-04-03 manuscript/manuscript.Rmd
Add keywords
Add FS suggestions

**Purpose:** To develop accelerometry-based equations to predict peak ground reaction forces (pGRF) and peak loading rate (pLR) during walking and running. **Methods:** `r Words(max(sample_size$pGRF_N))` subjects (`r unname(sample_size_sex["F"])` females; `r sample_descriptives$body_mass_mean`±`r sample_descriptives$body_mass_sd` kg) walked and ran at different speeds (2–14km·h^-1^) on a force plate–equipped treadmill while wearing accelerometers at ankle, lower back and hip. Regression equations were developed to predict pGRF and pLR from accelerometry data. Leave-one-out cross-validation was used to calculate prediction accuracy and Bland–Altman plots. Our pGRF prediction equation was compared with a reference equation previously published. **Results:** Body mass and peak acceleration were included for pGRF prediction and body mass and peak acceleration rate for pLR prediction. All pGRF equation coefficients of determination were above `r myround(max(GRF_accuracy$R2_cond), 2)`, and a good agreement between actual and predicted pGRF was observed, with a mean absolute percent error (MAPE) below `r myround(max(GRF_accuracy$MAPE) * 100, 1)`\%. Accuracy indices from our equations were better than previously developed equations, with a MAPE 2 to 3 times better. All pLR prediction equations presented a lower accuracy compared to those developed to predict pGRF. **Conclusion:** Walking and running pGRF can be predicted with high accuracy by accelerometry-based equations, representing an easy way to determine mechanical loading in clinical settings. The pLR prediction equations yielded an unsatisfactory accuracy.

@@

Exercise-related mechanical loading plays a key role in bone health, by promoting positive effects on bone mass, geometry and microarchitecture [@Turner2003; @Fonseca2014]. However, not all forms of mechanical stimulation are equally osteogenic, and the manipulation of variables such as loading magnitude, rate, frequency and volume can produce different bone responses [@Turner1998; @Turner2003]. Therefore, the correct assessment of mechanical loading, specially in a clinical environment, is of paramount importance to attain a significant positive bone response to exercise. Nonetheless, a limitation in this context is that the mechanical loading evaluation is usually performed only by instruments restrained to laboratory settings.

@@

Force plates (FP) are the golden standard instrument for mechanical loading measurement [@Medved2000]. However, their cost, need of expertise to collect and analyze data, of a dedicated indoor space and ongoing calibration and maintenance hinders its application in clinical settings. Accelerometers, due to their portability, low-cost and accuracy, are widely used to objectively measure physical activity related to cardiometabolic outcomes [@Troiano2014; @Migueles2017], and, more recently, have been proposed as adequate devices to predict gravitational loading variables [@Neugebauer2012; @Fortune2014; @Neugebauer2014; @Neugebauer2018; @Veras2020]. Given the need to assess mechanical loading in clincal settings in order to control and monitor exercise prescription for bone health, accurate mechanical loading prediction models need to be developed and validated for the most common activities prescribed to induce bone adaptations, such as walking, running, and jumping. These models, apart from being accurate, need to be transparent in their data processing methods, allowing replicability of the results, and ideally offer a tool to facilitate data processing and analysis for those who are not familiar with the employed techniques, such as what is already offered for analysing cardiometabolic variables [@Migueles2019], and thus spread their applicability.

@@

Therefore, given the necessity of an accurate and reliable method to assess mechanical loading in clinical settings in order to better monitor and prescribe exercise to obtain positive bone health outcomes, the aim of this study was to develop pGRF and pLR prediction models based on accelerometer data during walking and running.

@@

During the protocol, participants wore three activity monitors, one at their right hip (along the anterior axillary line, at the level of the iliac crest), another at their lower back (at the midpoint between the two posterior superior iliac spines) and the last at their right ankle (immediately superior to the lateral malleolus). Activity monitors at waist level were secured on the same elastic belt with a clip and the activity monitor placed on the ankle was secured by an elastic belt fixed with adhesive tape. In all placements the accelerometer vertical axis was aligned with the standing body longitudinal axis. All participants wore the same three activity monitors always positioned at the same places.

@@

To the best of our knowledge, only two studies have already attempted to develop pLR prediction models based on accelerometry data, and both studies only included walking activities. Accordingly to what we found, these studies also showed that the prediction error for these models is considerably larger than for the pGRF prediction. One of these researches investigated the use of four different accelerometer placements (ankle, tibia, thigh and waist) to predict pLR and achieved a MAPE ranging from 23.4\% to 46.1\% [@Fortune2014], the other one found values from 19.5\% to 22.4\% using accelerometers at the lower back and hip [@Veras2020]. Results from the models developed in both these investigations ranged from lower to higher prediction error compared to our results, which are approximately of 29.0\%. Although pLR is an important variable, known for a long time to affect the bone response to mechanical loading [@Bassey1998; @Turner1995], based on this and previous studies [@Fortune2014; @Veras2020], its prediction through accelerometer data does not seem to yield trustworthy results. Nevertheless, past research has suggested that the bone response to pGRF and pLR is somewhat similar due to the high correlation between these variables [@Heikkinen2007]. Therefore, the pGRF prediction by accelerometry data, which is shown to be accurate, already incorporates some of the LR effect on the bone response to loading.

>>>>

**Purpose:** To develop accelerometry-based equations to predict peak ground reaction forces (pGRF) and peak loading rate (pLR) during walking and running. **Methods:** `r Words(max(sample_size$pGRF_N))` subjects (`r unname(sample_size_sex["F"])` females; `r sample_descriptives$body_mass_mean`±`r sample_descriptives$body_mass_sd` kg) walked and ran at different speeds (2–14km·h^-1^) on a force plate–equipped treadmill while wearing accelerometers at ankle, lower back and hip. Regression equations were developed to predict pGRF and pLR from accelerometry data. Leave-one-out cross-validation was used to calculate prediction accuracy and Bland–Altman plots. Our pGRF prediction equation was compared with a reference equation previously published. **Results:** Body mass and peak acceleration were included for pGRF prediction and body mass and peak acceleration rate for pLR prediction. All pGRF equation coefficients of determination were above `r myround(max(GRF_accuracy$R2_cond), 2)`, and a good agreement between actual and predicted pGRF was observed, with a mean absolute percent error (MAPE) below `r myround(max(GRF_accuracy$MAPE) * 100, 1)`\%. Accuracy indices from our equations were better than previously developed equations, with a MAPE 2 to 3 times better. All pLR prediction equations presented a lower accuracy compared to those developed to predict pGRF. **Conclusion:** Walking and running pGRF can be predicted with high accuracy by accelerometry-based equations, representing an easy way to determine mechanical loading in free-living conditions. The pLR prediction equations yielded an unsatisfactory accuracy. **Key Words:** ACTIVITY MONITOR, FORCE PLATES, LOCOMOTION, RAW ACCELERATION, MECHANICAL LOADING.

@@

Physical activity-related mechanical loading plays a key role in bone health, by promoting positive effects on bone mass, geometry and microarchitecture [@Turner2003; @Fonseca2014]. However, not all forms of mechanical stimulation are equally osteogenic, and the manipulation of variables such as loading magnitude, rate, frequency and volume can produce different bone responses [@Turner1998; @Turner2003]. Therefore, the correct assessment of mechanical loading is of paramount importance to better understand the bone response to physical stimulus. Nonetheless, a limitation in this context is that the mechanical loading evaluation is usually performed only by instruments restrained to laboratory settings.

@@

Force plates (FP) are the golden standard instrument for mechanical loading measurement [@Medved2000]. However, their cost, need of expertise to collect and analyze data, of a dedicated indoor space and ongoing calibration and maintenance hinders its application in free-living conditions. Accelerometers, due to their portability, low-cost and reliability, are widely used to objectively measure physical activity related to cardiometabolic outcomes [@Troiano2014; @Migueles2017], and, more recently, have been proposed as adequate devices to predict gravitational loading variables [@Neugebauer2012; @Fortune2014; @Neugebauer2014; @Neugebauer2018; @Veras2020]. Given the need to assess mechanical loading in out-of-laboratory environments in order to control and monitor exercise prescription for bone health, accurate mechanical loading prediction models need to be developed and validated for the most common activities prescribed to induce bone adaptations, such as walking, running, and jumping. These models, apart from being accurate, need to be transparent in their data processing methods, allowing replicability of the results, and ideally offer a tool to facilitate data processing and analysis for those who are not familiar with the employed techniques, such as what is already offered for analysing cardiometabolic variables [@Migueles2019], and thus spread their applicability.

@@

Therefore, given the necessity of an accurate and reliable method to assess mechanical loading in free-living conditions in order to better monitor and prescribe exercise to obtain positive bone health outcomes, the aim of this study was to develop pGRF and pLR prediction models based on accelerometer data during walking and running.

@@

During the protocol, participants wore three activity monitors, one at their right hip (along the anterior axillary line, at the level of the iliac crest), another at their lower back (at the midpoint between the two posterior superior iliac spines) and the last at their right ankle (immediately superior to the lateral malleolus). Activity monitors at waist level were secured on the same elastic belt with a clip and the activity monitor placed on the ankle was secured by an elastic belt fixed with adhesive tape. In all placements the accelerometer vertical axis was aligned with the standing body longitudinal axis.

@@

To the best of our knowledge, only two studies have already attempted to develop pLR prediction models based on accelerometry data, and both studies only included walking activities. Accordingly to what we found, these studies also showed that the prediction error for these models is considerably larger than for the pGRF prediction. One of these researches investigated the use of four different accelerometer placements (ankle, tibia, thigh and waist) to predict pLR and achieved a MAPE ranging from 23.4\% to 46.1\% [@Fortune2014], the other one found values from 19.5\% to 22.4\% using accelerometers at the lower back and hip [@Veras2020]. Results from the models developed in these investigations have both lower and higher prediction errors compared to our results, which are approximately of 29.0\%. Although pLR is an important variable, known for a long time to affect the bone response to mechanical loading [@Bassey1998; @Turner1995], based on this and previous studies [@Fortune2014; @Veras2020], its prediction through accelerometer data does not seem to yield trustworthy results. Nevertheless, past research has suggested that the bone response to pGRF and pLR is somewhat similar due to the high correlation between these variables [@Heikkinen2007]. Therefore, the pGRF prediction by accelerometry data, which is shown to be accurate, already incorporates some of the LR effect on the bone response to loading.
